Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-13 Gandhi_Project_Report/draft4.tex
Changed 4.3 and H2. Conclusion writtend and added

% \begin{figure}[ht]
%   \includegraphics[scale=0.3]{_learningfailure.png}
%   \centering
%   \caption{Illustration of a system having failed to learn the map}
%   \label{fig:supp_learning_failure}
% \end{figure}

@@

\section {Unique Solution Property}

\begin{Definition}
  [\bf Unique Solution Property] \label{Dfn_usp}\rm
  A driven system $g$ is said to have the Unique Solution Property (USP) if for each input $\overline{u}$ there exists exactly one solution. 
  Alternatively we may formulate the USP as follows: $g$ has the Unique Solution Property if there exists a well-defined map $\Psi:{U}\to{X}$ with $\Psi({\overline{u}})$ denoting the unique solution.
\end{Definition}

One of the first result obtained upon defining the USP, is the fact that every solution will attract any different initial conditions towards the component parts of the solution. %, i.e. that the solution $\Psi$ is a non-autonomous uniform attractor. \ednote{Define Uniform Attractor before stating theorem or else omit the statement.} 
If $g$ has the USP, then any solution to $g$ is a uniform attractor in the nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this thesis and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

% \ednote{M: You can omit the theorem, and just say it is a uniform attractor in the nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}, and say the discussion on nonautonomous attractors is beyond the scope of the thesis.}
% Paraphrased.
 In our context, $g$ having the USP is equivalent to $g$ being a topological contraction \cite{manjunath2021universal}, or alternatively that $g$ exhibits the Echo State Property (ESP).

This property is used to describe the concept of fading memory or memory-loss in the majority of literature within the field reservoir computing (RC) \textbf{(Citation)}. 
$g$ having the ESP guarantees that the entire left-infinite history of an input will determine exactly the current system state; there is only ever the possibility of a single reachable state at a given instance of time \cite{jaeger2001echo}
The ESP is also often discussed in terms of a stability property that holds for all inputs of a system~\cite{manjunath2020stability} and, furthermore, plays a key role in the design and training of recurrent neural networks within the field of RC \cite{Manju_ESP}.

Defined formally, the ESP is intrinsically linked to an input (see \cite{Manju_ESP}) Unfortunately, definitions and results concerning the ESP do not have much practical use since a relationship of the network's response to the temporo-statistical properties of the input is not directly described. A more detailed presentation may be found in~\cite{jaeger2001echo}.

@@

Having already defined the reachable set $X_U$, we now pause for a moment to fix additional notation.
Letting $\cev{u}^{n}:=(\ldots,u_{n-2} ,u_{n-1})$ be the left-infinite subsequence of an input, $\overleftarrow{U}$ then denotes all these left-infinite sequences in $U$. 
Moreover, $\cev{u}^{n}v:=(\ldots,u_{n-2} ,u_{n-1}, v)$ is to symbolise the input up to time $n$ with $v \in U$ being the specific input value at time $n$. 
The introduction of a new input at time $n$ can be described by a mapping $\sigma_v:   \cev{u}^{n} \mapsto \cev{u}^{n}v$.

The question now becomes whether we may establish a semi-conjugacy %\ednote{M: I have changed it to a semi-conjugacy since conjugacy amounts to asking a infinite-dimensional space $\overleftarrow{U}$ to be embedded in $X_U$ which is not possible.} 
as presented below for the driven system $g$.

@@

We direct our attention to the function \[H_2{\overline{}} := (h(r\overline{u}, h(\overline{u}))) \]

An embedding of the space $\overleftarrow{U}$  in $\overleftarrow{U}$ would allow one to establish topological conjugacy, which in turn provides stronger results than merely obtaining a coarse-grained representation via a causal mapping \cite{M: I feel this is very confusing to the reader as the co-domain of causal mapping is the left-infinite space and , so please omit if possible. You can just say we want to restrict inputs to the left-infinite orbits of a dynamical system and then explore conjugacy instead of a mere semi-conjugacy}.

Before formalising this, we make mention of the concept of the inverse-limit system of a dynamical system $(U,T)$. 
These spaces are noteworthy as their topology can help one ascertain much about the original system \cite{Manju_IEEE}. The topological structure for inverse-limit systems of chaotic dynamical systems is complicated, so we relegate further discussion to \cite{kennedy2008inverse_limit, ingram2011inverse}.
In broad sweeps, the inverse-limit space of a dynamical system is a self-map on a subset of an infinite-dimensional space where each point in the inverse-limit space corresponds to a backward orbit of the map $T$. 
Following \cite{manjunath2021universal, ingram2011inverse}:

@@

\begin{Definition}
  [\bf Causal Embedding]\label{Dfn_CausalEmbed}\rm
  A driven system $g$ is said to causally embed the dynamical system $(U,T)$ if 
  \vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item The diagram in \ref{Scomm_h} commutes (i.e. $h$ is a universal semi-conjugacy),
  \item $H_2(\overline{u}):=(h(r\overline{u}, h(\overline{u})))$ embeds the inverse-limit space $(\widehat{U}_T, \hat{T})$ of $(U,T)$ in $X\times{X}$.
\end{enumerate}
\end{Definition}

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is also injective (in addition to being surjective, and hence bijective), 
it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$ and we refer to $H$ as a \emph{causal embedding} 
\ednote{M : Causal embedding a dynamical system is different from the definition of causal embedding, so please edit/delete this paragraph and the next theorem and also in the remaining part till Section 1.2}.

\begin{Theorem}
 If $g(\ldotp,x)$ is invertible and has the USP then $H$ is a causal embedding. 
\end{Theorem}
\begin{proof}
  See \cite{manjunath2021universal}.
\end{proof}

When $g$ has the USP, the diagram below illustrates the operation of the mappings $h$ and $H$. The mapping $h:\overleftarrow{U}\to{X_U}$ is an observable as discussed in an above section (\ref{TakensEmbedding}).

\begin{figure}[ht]
  %\includegraphics[scale=0.3]{actionofh_H.png}
  \centering
  \label{fig:actionh_H}
\end{figure}

The next result is of relative import but to fully comprehend its influence, some discussion still remains. We state it in part here.
% Before stating it, we define a concept which will directly be shown to be equivalent to the USP.

\begin{Theorem}
  The following statements are equivalent:
  \vspace{-8mm}
  \begin{enumerate}[noitemsep, label=\roman*.]
    \item $g$ has the USP. 
    \item $g$ is a topological contraction.
  \end{enumerate}
\end{Theorem}
% \begin{proof}
%   (postponed) 
% \end{proof}

@@

One must be careful to avoid a choice of $g$ which would  add complexity to the obtained solution.

When a causal embedding $H$ exists for the driven system $g$, one can map an arbitrary input ${u}$ onto the solution space $X$ without additional distortion or information-loss \textbf{(Cite.)}.
When an embedding is established, the question of possible additional complexity in the solution is removed by guaranteeing that, since the systems are conjugate (semi-conjugate, \textbf{(refer)}), $g$ does not add any (some) complexity to the system.

@@

Note that $H_2$ maps an entire left-infinite solution sequence from $\Psi$ to an element in $X\times{X}$.
We now have the following (compare with \ref{SCausal_H} above):

% \ednote{Add Causal Embedding Theorem}
% \textbf{NB NB NB - Causal Embedding Theorem! - Th.4}

\begin{equation} \label{Fig_inverse_limsystem}
  %  \[ 
      \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
      \everypsbox{\scriptstyle}
      \begin{psmatrix}
      \widehat{U}_T  && \widehat{U}_T \\
      Y_T && Y_T \\
      %%%
     %  \ncline{1,1}{1,2}^{\widehat{T}} \ncline{1,1}{2,1} <{h}
     %  \ncline{1,2}{2,2} > {H_2}
     %  \ncline{2,1}{2,2}^{G_T}
      \end{psmatrix}
    %  \]
  \end{equation}

@@

The graph in equation \ref{Fig_inverse_limsystem} is then extended as below:

% \begin{figure}[ht]
%   \includegraphics[scale=0.3]{_gammatemp.png}
%   \centering
%   \label{fig:temp_Gamma}
% \end{figure}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.35]{_dp_setup.png}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP. Image taken from~\cite{DPSetup}}
  \label{fig:dp_setup}
\end{figure}

The scalar equations of motion are derived using Newtonian physics and can be found in~\cite{DPFormulas}. The second-order system is converted to a first-order ODE and solved using the Runge-Kutta numerical method.  Noise-free, undamped DP data-sets were generated and utilised in this project due to the requirement that the map $T$ be surjective.
It may easily be verified that a DP undergoing damping (as seen in \ref{conjugacy_with_Gamma}) is not surjective and hence the required assumptions to apply our theory do not hold. \ednote{B: Change sentence.}

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dp_dying.eps}
  \centering
\caption{A damped pendulum is not surjective.}
\label{fig:damped_pendulum}
\end{figure}

\textbf{Parameter Values \& Successes}

Train length: 15000. Prediction Length: 1000. Discard to wait for network to forget: 500. Dimension of network: 128 neurons.
$\alpha=0.99$. $a=0.5$.

Hidden Layers: 16. Layer dimension: 64. Epochs:256. Batch Size: 128. Delay(for MDD): 8. Density of Sparse Matrices: 0.2. $d=1$

\ednote{B: Write this paragraph.}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{_DP_Fig-2F2023.eps} 
  \captionof{figure}{Predicted(red) and actual(blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads.} 
\end{figure}
\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{_DP_Fig-2B2023.eps}
  \captionof{figure}{Densities of predicted and actual values of x-coordinate of first pendulum head. }
  \end{figure}
  \begin{figure}[ht]
    \centering 
  \includegraphics[width=\linewidth]{_DP_Fig-2D_5.eps}
    \captionof{figure}{Takens Delay Map with delay=5 to illustrate behaviour of predicted and actual values of double pendulum data.}
\end{figure}

\subsection{Other attractors I’ve experimented with: Clifford \& Thomas}
We follow the methodology presented in \cite{manjunath2021universal} but opt to consider attractors different from the Lorenz system and H\'enon map in this project.

@@

The behaviour of the system with a $\beta$-parameter value of 0.1056 is simulated by scaling it to fit inside the interval $[-1,1]$ and both perturbed with noise as well as leaving it intact. Specifically the noise is obtained from a normally distribution with mean 0 and standard deviation 0.05, which translates approximately to *dB.
 A sequence of 3000 observations in the clear and noisy data-set respectively is fed into the system with 300 entries discarded before predicting the next 9000 entries.

\begin{figure}[ht]
  \centering
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{Thomas_nonoise.eps}
    \caption*{True and predicted trajectories in the absence of noise.}
  \endminipage\hfill
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{ThomasNoisy.eps}
    \caption*{True and predicted trajectories for noise added.}
  \endminipage
\end{figure}

The RNN was initialised as follows: The dimension of the network is set equal to 500 neurons and 12 hidden layers of dimension 64 each were initialised. Training is realised through 150 epochs of batch sizes of 128 each; the map is learnt via the Adam Optimiser using finer and finer learning rates (0.001 and then divided by 10 at every iteration).

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.35]{Thomas_1.eps}\caption*{Predicted trajectories for the attractor demonstrate empirically the ability to predict the evolution of the trajectory for the next ~100 timesteps}
  \includegraphics[scale=0.5]{Thomas_3.eps}\caption*{Represented here the learnt (blue) and actual (red) densities of the first coordinate of the dataset.}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth,left]{Clifford_1.eps}
  \caption*{These graphs were constructed by predicting the Clifford system 15000 steps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. As perceived here, we are able to lock on to the trajectory of the Clifford map almost exactly for the first 25 steps.}
  \minipage{0.5\textwidth}
      \centering
      \includegraphics[width=\linewidth]{Clifford_2.eps}
  \endminipage\hfill
  \minipage{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{Clifford_3.eps}
  \endminipage
  \caption{The bottom row, left, shows a plot of the true attractor(red) and the predicted values(blue). Bottom row, right, shows the distribution of the first coordinate of the learnt system (blue) and the actual Clifford system (red). }
  \label{Fig_Clifford}
\end{figure}

@@

      \caption{Following \cite{manjunath2021universal}, we determine the multidimensional correlation coefficients $\rho$ to evidence a general reduction in the functional complexity of the map $G_T$ emerging due to a RNN. Rows correspond to distinct dynamical systems considered in the experiments
      The second column corresponds to the amount of artificial neurons (the dimension of $X$) in the implemented RNN.
      The last two columns comprise numerical estimates of $\rho$ for the relevant vectors. The linear association between $u_n$ vs $T(u_{n})$, as well as the linear relationship $\begin{bmatrix}
              x_{n-1}\\
              x_n
          \end{bmatrix}$ vs $G_T \left( \begin{bmatrix}
              x_{n-1}\\
              x_n
          \end{bmatrix}\right)$ may be contrasted by the reader. 
          Finally, the RNN's dimension (from \eqref{Seq_RNN}) is provided to exhibiting that an increase in dimension of the driven system will 
          typically result in a map $G_T$ with less functional complexity.} 
          \label{Table_FC}
      \end{table}\label{tbl_attractorsPearson}
\end{center}
\ednote{B: Functional Complexity will be explained before this}

@@

The driven states $(x_{n-1},x_n)$ holds the information of the left-infinite history of the input. Furthermore, there is some evidence that the inverse-limit space (not system) seems to distinguish two dynamical systems that are not topologically conjugate for a class of systems in the sense their inverse-limit spaces are not homeomorphic to each other.  So the relation $Y_T$, which is homeomorphic to the inverse-limit system, may be unique to a class of systems. (The reference is a paper on Ingram’s Conjecture by Marcy Barge, Henk Bruin and Sonja Stimac.) Further investigation is needed.

\ednote{B: Expand}

>>>>

\ednote{Include note on learning failure because of sensitivity as in graph commented out?}
% \begin{figure}[ht]
%   \includegraphics[scale=0.3]{_learningfailure.eps}
%   \centering
%   \caption{Illustration of a system having failed to learn the map}
%   \label{fig:supp_learning_failure}
% \end{figure}

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of this requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of $U$, we me conclude that $T$ is not surjective as there exists some point in $U-A$ which does not have a preimage.
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the oscillations dying down for a damped pendulum. (A more in-depth account will be given in \ref{ch5})

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{I rememeber you saying this in one of our sessions, but don't have a source for it.}

@@

\section{Unique Solution Property}

\begin{Definition}
  [\bf Unique Solution Property] \label{Dfn_usp}\rm
  A driven system $g$ is said to have the Unique Solution Property (USP) if for each input $\overline{u}$ there exists exactly one solution. 
  Alternatively we may formulate the USP as follows: $g$ has the Unique Solution Property if there exists a well-defined map $\Psi:{U}\to{X}$, with $\Psi({\overline{u}})$ denoting the unique solution.
\end{Definition}

One of the first result obtained after defining the USP is the fact that every solution will attract different initial conditions towards the component parts of the solution. %, i.e. that the solution $\Psi$ is a non-autonomous uniform attractor. \ednote{Define Uniform Attractor before stating theorem or else omit the statement.} 
If $g$ has the USP, then any solution to $g$ is also a uniform attractor in nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this project and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

% \ednote{M: You can omit the theorem, and just say it is a uniform attractor in the nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}, and say the discussion on nonautonomous attractors is beyond the scope of the thesis.}
% Paraphrased.
Additionally, $g$ having the USP is in our context also equivalent to $g$ being a topological contraction \cite{manjunath2021universal}, or alternatively also that $g$ exhibits the Echo State Property (ESP).

In the majority of Reservoir Computing(RC) literature (a machine learning algorithm approach)\ednote{How should I explain/qualify RC better here? It's mentioned without really explaining.}, 
the ESP is used to describe the concept of fading memory or memory-loss  \cite{jaeger2004harnessing, lu2018attractor}. 
If $g$ possesses the ESP (equivalent in our context to the USP), the we are guaranteed that the whole of an input's left-infinite history will exactly determine the current system state; i.e. there is only ever the possibility of a single reachable state at a given instance of time \cite{jaeger2001echo}

Defined formally, the ESP is intrinsically linked to an input (see \cite{Manju_ESP}) Unfortunately, definitions and results concerning the ESP do not have much practical benefit since a relationship between the network's response and the temporo-statistical properties of the input are not directly described. A more detailed presentation may be found in~\cite{jaeger2001echo}.

We do however consider it important to mention that the ESP is also often discussed in terms of a stability property that holds for all inputs of a system~\cite{manjunath2020stability} and, furthermore, plays a key role in the design and training of recurrent neural networks(RNN) within the field of RC \cite{Manju_ESP}.

@@

Having already defined the reachable set $X_U$ as the set of all elements of all entire solutions, we pause briefly in order to fix additional notation.
Defining $\cev{u}^{n}:=(\ldots,u_{n-2} ,u_{n-1})$ as the left-infinite subsequence of an input up until time $n$, $\overleftarrow{U}$ is then the notation for all these left-infinite sequences in $U$. 
Moreover, $\cev{u}^{n}v:=(\ldots,u_{n-2} ,u_{n-1}, v)$ will symbolise the input up to time $n$ with $v \in U$ being the specific input value at time $n$. 
The introduction of a new input at time $n$ can be described by the mapping $\sigma_v:   \cev{u}^{n} \mapsto \cev{u}^{n}v$. 
The right-shift map, $r\overleftarrow{U}\to\overleftarrow{U}$, of an input sequence is defined $r: (\cdots, u_{-2},u_{-1}) \mapsto(\cdots, u_{-3},u_{-2})$

The goal now assumes the form of the question on our ability to establish a semi-conjugacy as presented below for the driven system $g$. 
%\ednote{M: I have changed it to a semi-conjugacy since conjugacy amounts to asking a infinite-dimensional space $\overleftarrow{U}$ to be embedded in $X_U$ which is not possible.}

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. \ednote{Explain why $H_{-1}$ is continuous} and we refer to $H$ as a \emph{causal embedding} 
% \ednote{M : Causal embedding a dynamical system is different from the definition of causal embedding, so please edit/delete this paragraph and the next theorem and also in the remaining part till Section 1.2}. 
% \begin{Theorem}
%  If $g(\ldotp,x)$ is invertible and has the USP then $H$ is a causal embedding. 
% \end{Theorem}
% \begin{proof}
%   See \cite{manjunath2021universal}.
% \end{proof}
When $g$ has the USP, the diagram below (adopted from~\cite{Manju_Nonlinearity}) illustrates the operation of the mappings $h$ and $H$. The mapping $h:\overleftarrow{U}\to{X_U}$ is also considered an observable as mentioned in the introduction of Chapter \ref{ch3}).

\begin{figure}[ht]
  \includegraphics[scale=0.3]{_actionofh_H.eps}
  \centering
  \captionof{figure}{Ladder-like behaviour of $h$ and $H$ to illustrate the causal mapping. }
  \label{fig:actionh_H}
\end{figure}

We wish to explore the possibility of establishing a conjugacy rather than a mere semi-conjugacy and therefore are interested in the restriction of inputs to the left-infinite orbits of a dynamical system.
The left-infinite orbit spaces are significant because a great amount of information about the original system can be gleaned from their topological structure~\cite{Manju_IEEE}. 
The formal topological structure for the so-called inverse-limit space of chaotic dynamical systems is complicated, and further discussion is relegated to~\cite{kennedy2008inverse_limit, ingram2011inverse}.
Broadly, we may conceive of the inverse-limit space of a dynamical system as a self-map on a subset of an infinite-dimensional space where each point in the inverse-limit space corresponds to a backward orbit of the map $T$. 
Following~\cite{manjunath2021universal, ingram2011inverse} we provide a definition:

@@

The inverse-limit space is well-defined since T : U →U is surjective by assumption.

@@

The driving function $g$ cannot be chosen in a frivolous manner as it could possibly complicate our work. Specifically, one must be careful to avoid a choice of $g$ which would  add complexity to the obtained solution.

When a causal embedding $H$ exists for the driven system $g$, one can map an arbitrary input ${u}$ onto the solution space $X$ without additional distortion or information-loss \textbf{(Cite.)}.
Established such an embedding eliminates the question of possible additional complexity in the solution by guaranteeing that, since the systems are conjugate (semi-conjugate, \textbf{(refer)}), $g$ does not add any (some) complexity to the system.

@@

Recall that $H_2$ maps an entire left-infinite solution sequence from $\Psi$ to an element in $X\times{X}$.
We now have the following (which may be compared with \ref{SCausal_H} above):

\begin{equation} \label{fig:inverse_limsystem}
  %  \[ 
      \psset{arrows=->, arrowinset=0.25, linewidth=0.6pt, nodesep=3pt, labelsep=2pt, rowsep=0.7cm, colsep = 1.1cm, shortput =tablr}
      \everypsbox{\scriptstyle}
      \begin{psmatrix}
      \widehat{U}_T  && \widehat{U}_T \\
      Y_T && Y_T \\
      %%%
     %  \ncline{1,1}{1,2}^{\widehat{T}} \ncline{1,1}{2,1} <{h}
     %  \ncline{1,2}{2,2} > {H_2}
     %  \ncline{2,1}{2,2}^{G_T}
      \end{psmatrix}
    %  \]
  \end{equation}

@@

The graph in equation \ref{fig:inverse_limsystem} is then extended as below:

% \begin{figure}[ht]
%   \includegraphics[scale=0.3]{_gammatemp.eps}
%   \centering
%   \label{fig:temp_Gamma}
% \end{figure}

@@

\begin{figure}[ht]
  \includegraphics[scale=0.35]{_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP. Image taken from~\cite{DPSetup}}
  \label{fig:dp_setup}
\end{figure}

The scalar equations of motion are derived using Newtonian physics and can be found in~\cite{DPFormulas}. The second-order system is converted to a first-order ODE and solved using the Runge-Kutta numerical method.  Noise-free, undamped DP data-sets were generated and utilised in this project due to the requirement that the map $T$ be surjective. We discuss the case of a DP exhibiting energy-loss (i.e. where the map $T$ is not surjective) below.

\textbf{Data originating from a non-surjective map.}

The importance of the requirement that data originates from a surjective map is observed learning failure of a map describing a damped DP.
Due to the assumption, the program discounted any possibility of energy-loss and we consider the example below.

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dp_dying.eps}
  \centering
\caption{The map generating data for a damped pendulum is not surjective.}
\label{fig:damped_pendulum}
\end{figure}

Despite feeding data that originated from a DP with dying occilations into the system, a map was instead learnt which continued to oscillate in a chaotic, but conservative manner.

\begin{figure}[ht]
  \includegraphics[scale=0.4]{_dpfailure_nonsurj.eps}
  \centering
\caption{Failure to learn a map which is not surjective.}
\label{fig:dp_notsurjective}
\end{figure}

Having spent ample time to discuss the challenges, we now change tone so as to satisfy the optimist.
We present successful run for DP-data to repdict the future movement of the pendulum.

The system was driven for 15500 steps, of which the first 500 were immediately discarded to simulate the network 'forgetting'. The driving function $g$ had input and reservoir matrices were implemented as sparse matrices filled to 20\%. Parameter-values of $\alpha=0.99$ and $a=0.5$ were used.
The remaining 15000 datapoints were used to train a network with 128 neurons. 
The network was initialised to contain 16 hidden layers each with a layer dimension of 64, activated with the tanh activation function.
Training was accomplished using 256 epochs with batch sizes 128 and a delay of 1.

% Trainlen:15000; Predlen: 1000; Discard: 500; Dimension:128 neurons; $\alpha=0.99$; $a=0.5$. Hidden Layers: 16. Layer dimension: 64. Epochs:256. Batch Size: 128. Delay(for MDD): 8. Density of Sparse Matrices: 0.2. $d=1$

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{_DP_Fig-2F2023.eps} 
  \captionof*{figure}{Predicted(red) and actual(blue) trajectories of the $x$- and $y$-coordinates of each of the pendulum heads.} 
% \end{figure}
% \begin{figure}[ht]
%   \centering
  \includegraphics[width=0.8\linewidth]{_DP_Fig-2B2023.eps}
  \captionof*{figure}{Densities of predicted and actual values of x-coordinate of first pendulum head. }
  % \end{figure}
  % \begin{figure}[ht]
  %   \centering 
  \includegraphics[width=0.8\linewidth]{_DP_Fig-2D_5.eps}
    \captionof*{figure}{Takens Delay Map with delay=5 to illustrate behaviour of predicted and actual values of double pendulum data.}
    \label{fig:dp_predictions}
\end{figure}

\subsection{Additional Attractors: Clifford \& Thomas}
We follow the methodology presented in \cite{manjunath2021universal} but opt to consider attractors different from the Lorenz system and H\'enon map in this project.

@@

The behaviour of the system with a $\beta$-parameter value of 0.1056 was simulated by scaling it to fit inside the interval $[-1,1]$. Two cases were considered: noise-free and perturbed. Noise was generated from a normal distribution with mean 0 and standard deviation 0.05, which translates approximately to *dB.
 A sequence of 3000 observations in both the clear and noisy data-sets respectively were fed into the system with 300 entries discarded to simulate the network's loss of memory. The next 9000 steps were predicted and compared with true data.

\begin{figure}[ht]
  \centering
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{_Thomas_nonoise.eps}
    \caption*{True and predicted trajectories in the absence of noise.}
  \endminipage\hfill
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{_ThomasNoisy.eps}
    \caption*{True and predicted trajectories for noise added.}
  \endminipage
\end{figure}

The RNN was initialised as follows: the dimension of the network is set equal to 500 neurons and 12 hidden layers of dimension 64 each were initialised. Training is realised through 150 epochs of batch sizes of 128 each; the map is learnt via the Adam Optimiser using finer and finer learning rates (0.001 and then divided by 10 at every iteration).

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.35]{_Thomas_1.eps}\caption*{Predicted trajectories for the attractor demonstrate empirically the ability to predict the evolution of the trajectory for the next ~100 timesteps}
  \includegraphics[scale=0.5]{_Thomas_3.eps}\caption*{Represented here the learnt (blue) and actual (red) densities of the first coordinate of the dataset.}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth,left]{Clifford_1.eps}
  \caption*{These graphs were constructed by predicting the Clifford system 15000 steps into the future and in so doing illustrating the long-term consistency and accuracy of the learnt system. As perceived here, we are able to lock on to the trajectory of the Clifford map almost exactly for the first 25 steps.}
  \minipage{0.5\textwidth}
      \centering
      \includegraphics[width=\linewidth]{Clifford_2.eps}
  \endminipage\hfill
  \minipage{0.5\textwidth}
    \centering
    \includegraphics[width=\linewidth]{Clifford_3.eps}
  \endminipage
  \caption{The bottom row, left, shows a plot of the true attractor(red) and the predicted values(blue). Bottom row, right, shows the distribution of the first coordinate of the learnt system (blue) and the actual Clifford system (red). }
  \label{fig:Clifford}
\end{figure}

@@

      \caption{Following \cite{manjunath2021universal}, we determine the multidimensional correlation coefficients $\rho$ to evidence a general reduction in the functional complexity of the map $G_T$ emerging due to a RNN. Rows correspond to distinct dynamical systems considered in the experiments
      The second column corresponds to the amount of artificial neurons (the dimension of $X$) in the implemented RNN.
      The last two columns comprise numerical estimates of $\rho$ for the relevant vectors. The linear association between $u_n$ vs $T(u_{n})$, as well as the linear relationship $\begin{bmatrix}
              x_{n-1}\\
              x_n
          \end{bmatrix}$ vs $G_T \left( \begin{bmatrix}
              x_{n-1}\\
              x_n
          \end{bmatrix}\right)$ may be contrasted by the reader. 
          Finally, the RNN's dimension (from \eqref{Seq_RNN}) is provided to exhibiting that an increase in dimension of the driven system will 
          typically result in a map $G_T$ with less functional complexity.} 
          \label{Table_FC}
      \end{table}\label{tbl_attractorsPearson}
\end{center}

@@

% The driven states $(x_{n-1},x_n)$ hold the information of the left-infinite history of the input. Furthermore, there is some evidence that the inverse-limit space (not system) seems to distinguish two dynamical systems that are not topologically conjugate for a class of systems in the sense their inverse-limit spaces are not homeomorphic to each other.  So the relation $Y_T$, which is homeomorphic to the inverse-limit system, may be unique to a class of systems. (The reference is a paper on Ingram’s Conjecture by Marcy Barge, Henk Bruin and Sonja Stimac.) Further investigation is needed.  \cite{budivsic2012applied, korda2020data}

As technology has advanced over the past century, humanity’s ability to record and manipulate rich, complex data in order to produce forecasts has profoundly increased. 
Collecting measurements from underlying artificial, natural, and physical systems through observables to realise a dynamical model with greater descriptive power is well-established in present-day science.

A key to the success of these techniques is the fact that dynamics discerned from the observed space can often be characterised by a mapping of much lower functional complexity when compared to the map describing the unknown underlying system. Takens theorem and the numerous generalizations (e.g., [5, 6, 7, 8, 8]) thereof is one such method establishing (under some generic conditions) the learnability of a system produced by creating a delay-coordinate vector from a sufficiently large number of previous observations of a dynamical system. This system is topologically conjugate to the system from which the observed time-series was obtained.

A critical concern of this method is the requirement of full knowledge of the variables of the underlying state space that govern the complex dynamical system one is interested in. In real-world conditions this intelligence is seldom available and one must depend heavily on human expertise and insight.

Each discrete-time state space model has at its core a driven dynamical system. In this project, it has been shown that an input originating from a compact space entailing exactly one solution (i.e., if the driven system has the USP) is equivalent to representing information without distortion, a concept which was formalised by establishing a causal mapping’s existence. The established results are general, implying that any driven dynamical system with specific properties can give rise to a causal mapping, and in so doing  providing an answer as to the success of arbitrary choices of driven systems in information processing applications in [4, 5].

In particular, we have a topological conjugacy (semi-conjugacy) between the data comprising the 11single-delay lag dynamics of a driven system’s and the underlying homeomorphic dynamical system. Finite-length input data is sufficient for forecasting as driven system states are influenced only in an almost negligibly small manner by the left-infinite segment of the past of the input beyond some arbitrarily large but finite time (a result which holds due to the universal semi-conjugacy of a driven system’s continuity)

Viewed through the lens reservoir computing approaches, we have also demonstrated that when single-delay dynamics are utilised, a learnable map in a reservoir computing network certainly exists. Moreover, such a driven system within a reservoir computing network enjoys certain proven qualities pertaining to the robustness w.r.t input- and parameter-perturbation. Accordingly, we are able to preserve the quality of a representation that would otherwise remain sensitive to various fluctuations, rendering the learning of the system’s dynamics wholly unpredictable. In addition, we obtain the robustness of the system against external noise, a phenomenon which is observed for the Fractal Dream Attractor.

Finally, we consider some physical systems to observe the methodology in practice and verify that we can indeed predict the future evolution of chaotic systems such as the double pendulum. Empirical evidence shows that a linearity measure like a Pearson correlation coefficient points at reduced functional complexity while learning the dynamics in the state space of the driven system

Considered with a philosophical-scientific lens, this project has aimed to establish a juncture between rationalism (the theory of nonautonomous dynamical systems) and empiricism (machine learning) to yield results profitable to the discipline of Applied Mathematics. This was done by showing that dynamics in an observed space can be established as topologically conjugate to the underlying system. As a result of the topological conjugacy, deep learning techniques are then subsequently applied to physically such equations from the data. In so doing, we acquire models of high fidelity that provide exceptional forecasting results on chaotic data that exhibit both statistical and topological consistency in the long run.
